---
title: "ps5-assignment"
author: "Rosemary Randall"
date: "`r Sys.Date()`"
output: html_document
---
## Loading and checking the data
```{r echo=FALSE}
library("tidyverse")
gaps <- read_delim("gapminder.csv.bz2")

row <- nrow(gaps)
col <- ncol(gaps)
```
There are `r row` rows in this dataset, attributed by `r col` columns. Here's a sample:
```{r}
gaps %>% 
  sample_n(10)
```
## Descriptive Statistics

```{r, echo=FALSE}
countName <- length(unique(gaps$name))
  
countIso3 <- length(unique(gaps$iso3))

countIso2 <- length(unique(gaps$iso2))
```
There are `r countIso3` 3-letter country name abbreviations, `r countName` unique country names, and `r countIso2` 2-letter country name abbreviations.

What's going on here? If there are fewer than 250 ISO2 abbreviations, that means that some are used multiple times. 
```{r}
gaps %>% 
  group_by(iso2) %>% 
  summarize(num = n_distinct(name)) %>% 
  arrange(desc(num)) %>% 
  head(3)
```
What about for the ISO3 abbreviations? Do any countries use multiple codes?
```{r}
gaps %>% 
  group_by(name) %>% 
  summarize(abbNum = n_distinct(iso3)) %>% 
  arrange(desc(abbNum)) %>% 
  head(3)
```
What are the most and least recent years in the data?
```{r}
max(gaps$time, na.rm = TRUE)

min(gaps$time, na.rm = TRUE)
```

## CO2 Emissions

How many missing CO2 emissions are there per year? Which years have the most missing?

```{r}
gaps %>% 
  group_by(gaps$time) %>% 
  filter(is.na(co2) | is.na(co2_PC)) %>% 
  count() %>% 
  arrange(desc(n)) %>% 
  head(3)
```
What are the total CO2 emissions over time for the U.S., China, and India, plus additional countries?
```{r}
gaps %>% 
  filter(iso3 == c("USA", "IND", "CHN", "DNK", "COL", "EGY")) %>% 
  ggplot() + 
  geom_line(aes(x = time, y = co2)) + 
  geom_point(aes(x = time, y = co2))
```

Does this change when viewing CO2 emmittance on a *per capita* basis?
```{r}
gaps %>% 
  filter(iso3 == c("USA", "IND", "CHN", "DNK", "COL", "EGY")) %>% 
  ggplot() + 
  geom_line(aes(x = time, y = co2_PC)) + 
  geom_point(aes(x = time, y = co2_PC))
```